A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-16 paper/fascd.tex
clean appendix

@@

FIXME smoothers including Newton active set \cite{BensonMunson2006} or Newton semi-smooth \cite{BensonMunson2006}, with either direct linear solve or a fixed number of Krylov steps

@@

FIXME show results for 2D Firedrake implementation of FASCD for $p$-Laplacian obstacle problem (Example \ref{ex:plaplacian}), advection-diffusion obstacle problem (Example \ref{ex:advectiondiffusion}), and porous-media obstacle problem (Example \ref{ex:porous})

@@

FIXME constrast the implemented examples in \cite{GraeserKornhuber2009,Tai2003} in which $f$ is linear

@@

\appendix
\section{Reductions of FASCD to familiar multilevel algorithms} \label{app:reductions}

\pr{fascd-vcycle} (Algorithm \ref{alg:fascd}) in the main text generalizes two better-known multilevel algorithms.  First we show that it reduces to the FAS multigrid V-cycle for nonlinear PDEs \cite{Trottenbergetal2001}, which itself generalizes linear multigrid, when inequality constraints are removed.  Second we reduce it to an existing V-cycle for unilateral linear variational inequalities \cite{GraeserKornhuber2009} in the case that the operator is linear and local.  Thus this Appendix separates FAS/nonlinear aspects of the algorithm from those modifications of linear multigrid which are caused by the inequality constraints.

To derive the classical FAS multigrid V-cycle we remove the inequality constraints from the finest-level VI problem \eqref{eq:fe:vi} and then simplify the \pr{fascd-vcycle} pseudocode accordingly.  In fact, suppose that the finest-level obstacles $\underline{\gamma}^J$ and $\overline{\gamma}^J$ are very negative and positive, respectively, thus not constraining the solution $u^J$ or any possible iterate.  The admissible set $\mathcal{K}^J$ of \eqref{eq:fe:vi} is then determined only by the Dirichlet boundary conditions.  Let $\mathcal{V}_0^J \subset \mathcal{V}^J$ denote the linear subspace with zero values on $\partial_D\Omega$.  By replacing the (now) arbitrary element $v-u^J\in\mathcal{V}_0^J$ in \eqref{eq:fe:vi} with ``$v$'' we have the finest-level problem
\begin{equation}
\ip{f^J(u^J)}{v} = \ip{\ell^J}{v} \qquad \text{for all } v\in \mathcal{V}_0^J, \label{eq:app:fas:pde}
\end{equation}
the FE discretization of strong-form PDE $f(u)=\ell$.  The downward $\mathcal{D}^j$ and upward $\mathcal{U}^j$ constraint sets can also be enlarged so that the corrections $y^j$ and $z^j$ in \pr{fascd-vcycle} are never constrained.  Formally we might do this by choosing constant LDCs with $\underline{\chi}^j \ll 0 \ll \overline{\chi}^j$, but so that ordering property \eqref{eq:fe:chiordering} still holds, and also with large gaps so that $\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1} \ll 0 \ll \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1}$.  Thus the corrections $y^j$ and $z^j$ are unconstrained, and lines 2--3 and 5--8 can be removed from Algorithm \ref{alg:fascd}.

Now the downward/upward VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} are also unconstrained, and can be rewritten as weak forms $\ip{f^j(w^j + y^j)}{v} = \ip{\ell^j}{v}$ and $\ip{f^j(w^j + z^j)}{v} = \ip{\ell^j}{v}$, respectively, for all $v\in \mathcal{V}_0^j$.  However, we may simply refer to the unconstrained, approximate solution $u^j=w^j+y^j$, instead of $w^j$ and $y^j$ separately.  The smoothers and the coarsest-level solver can be regarded as unconstrained, in-place modifications of $u^j$, so we denote them simply as ``$\text{\pr{smooth}}(\ell^j;u^j)$.''  Noting line 11 of Algorithm \ref{alg:fascd} says $u^{j-1}=\iR u^j$, after down-smoothing the coarsened source functional computed by line 12 can be rewritten as
\begin{equation}
\ell^{j-1} = f^{j-1}\left(u^{j-1}\right) + R\left(\ell^j-f^j(u^j)\right). \label{eq:app:fas:levelsource}
\end{equation}

Going upward we must be careful that only the coarse correction, and not the coarse solution, is prolonged.\footnote{Compare Remark 5.3.9 in \cite{Trottenbergetal2001}, a warning to novices.}  After up-smoothing on the $j-1$ level the coarse iterate is $u^{j-1}$, but at this point $u^j=w^j + y^j$ contains the result of down-smoothing on the $j$th level.  On the other hand, in Algorithm \ref{alg:fascd} the initial value of the correction $z^j$, before up-smoothing on the $j$th level, is $z^j = y^j + P z^{j-1}$, where $z^{j-1}$ contains the total correction from coarser corrections and up-smoothing at the $j-1$ level.  Thus the $j$th-level iterate \emph{before} up-smoothing must be set to
    $$w^j + y^j + P z^{j-1} = u^j + P(u^{j-1} - \iR u^j),$$
a standard FAS assignment.

With the given modifications, we have the much-simplified pseudocode in Algorithm \ref{alg:fas}.  This FAS V-cycle algorithm appears in various places, for example Algorithm 14 of \cite{Bruneetal2015} or in section 5.3.4 of \cite{Trottenbergetal2001}.

\begin{pseudofloat}[ht]
\begin{pseudo} \label{ps:fas-vcycle}
\pr{fas-vcycle}(\ell^J; u^J)\text{:} \\+
    for $j=J$ downto $j=1$ \\+
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j;u^j)$ \\
      $u^{j-1} = \iR u^j$ \\
      $\ell^{j-1} = f^{j-1}(u^{j-1}) + R \left(\ell^j - f^j(u^j)\right)$ \\-
    $\text{\pr{solve}}(\ell^0;u^0)$ \\
    for $j=1$ to $j=J$ \\+
      $u^j \gets u^j + P (u^{j-1} - \iR u^j)$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j;u^j)$ \\-
\end{pseudo}
\caption{Removing inequality constraints from \pr{fascd-vcycle} yields the standard FAS V-cycle.}
\label{alg:fas}
\end{pseudofloat}

FIXME by assuming $f$ is linear, reduce to Algorithm 4.7 in \cite{GraeserKornhuber2009} but with $\text{V}(\nu_1,\nu_2)$ cycles

@@

FIXME move this from this appendix, after using it in the discussion etc.

>>>>

FIXME rewrite the entire introduction

@@

FIXME smoothers including Newton active set \cite{BensonMunson2006} or Newton semi-smooth \cite{BensonMunson2006}, with either direct linear solve or a fixed number of Krylov steps; in latter case either fixed number of CG+ICC or fixed number of GMRES+ILU

@@

FIXME show results for Firedrake implementation of FASCD for the 2D classical obstacle problem (ball and spiral), 1D $p$-Laplacian obstacle problem (Example \ref{ex:plaplacian}), advection-diffusion obstacle problem (Example \ref{ex:advectiondiffusion}), and SIA obstacle problem

@@

@@

\appendix
\section{Reduction of FASCD to FAS multigrid for PDEs} \label{app:reductions}

\pr{fascd-vcycle} (Algorithm \ref{alg:fascd}) in the main text generalizes existing multilevel algorithms.  Here we show that it reduces to the FAS multigrid V-cycle for nonlinear PDEs \cite{Trottenbergetal2001}, which itself generalizes linear multigrid, when all inequality constraints are removed.

Suppose we remove the inequality constraints from the finest-level VI problem \eqref{eq:fe:vi} by replacing the finest-level obstacles $\underline{\gamma}^J$ and $\overline{\gamma}^J$ with very negative and positive constants, respectively, thus not constraining the solution $u^J$ or any possible iterate.  The admissible set $\mathcal{K}^J$ of \eqref{eq:fe:vi} is then determined only by the Dirichlet boundary conditions.  Let $\mathcal{V}_0^J \subset \mathcal{V}^J$ denote the linear subspace with zero values on $\partial_D\Omega$.  By replacing the (now) arbitrary element $v-u^J\in\mathcal{V}_0^J$ in \eqref{eq:fe:vi} with ``$v$'' we have the finest-level problem
\begin{equation}
\ip{f^J(u^J)}{v} = \ip{\ell^J}{v} \qquad \text{for all } v\in \mathcal{V}_0^J, \label{eq:app:fas:pde}
\end{equation}
the FE discretization of strong-form PDE $f(u)=\ell$.  The downward $\mathcal{D}^j$ and upward $\mathcal{U}^j$ constraint sets can also be enlarged so that the corrections $y^j$ and $z^j$ in \pr{fascd-vcycle} are never constrained.\footnote{Formally we might do this by choosing constant LDCs with $\underline{\chi}^j \ll 0 \ll \overline{\chi}^j$, but so that ordering property \eqref{eq:fe:chiordering} still holds, and also with large gaps so that $\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1} \ll 0 \ll \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1}$.}  Thus the corrections $y^j$ and $z^j$ are unconstrained, and lines 2--3 and 5--8 can be removed from Algorithm \ref{alg:fascd}.

Now the downward/upward VIs \eqref{eq:fe:downvi} and \eqref{eq:fe:upvi} are also unconstrained, and can be rewritten as weak forms $\ip{f^j(w^j + y^j)}{v} = \ip{\ell^j}{v}$ and $\ip{f^j(w^j + z^j)}{v} = \ip{\ell^j}{v}$, respectively, for all $v\in \mathcal{V}_0^j$.  However, we may simply store the unconstrained, approximate solution $u^j=w^j+y^j$, instead of $w^j$ and $y^j$ separately, and similarly for the upward solution $u^j=w^j+z^j$.  The smoothers and the coarsest-level solver can be regarded as unconstrained, in-place modifications of $u^j$, so we denote them simply as ``$\text{\pr{smooth}}(\ell^j;u^j)$.''  Noting line 11 of Algorithm \ref{alg:fascd} now says $u^{j-1}=\iR u^j$, after down-smoothing the coarsened source functional computed by line 12 can be rewritten as
\begin{equation}
\ell^{j-1} = f^{j-1}\left(u^{j-1}\right) + R\left(\ell^j-f^j(u^j)\right). \label{eq:app:fas:levelsource}
\end{equation}

Going upward we must be careful that only the coarse correction, and not the coarse solution, is prolonged.\footnote{Compare Remark 5.3.9 in \cite{Trottenbergetal2001}, a warning to novices and everyone else.}  After up-smoothing on the $j-1$ level the coarse iterate is $u^{j-1}$, but at this point $u^j=w^j + y^j$ contains the result of down-smoothing on the $j$th level.  On the other hand, in Algorithm \ref{alg:fascd} the initial value of the correction $z^j$, before up-smoothing on the $j$th level, is $z^j = y^j + P z^{j-1}$, where $z^{j-1}$ contains the total correction from coarser corrections and up-smoothing at the $j-1$ level.  Thus the $j$th-level iterate \emph{before} up-smoothing must be set to
    $$w^j + y^j + P z^{j-1} = u^j + P(u^{j-1} - \iR u^j).$$

With these modifications, we have a much-simplified pseudocode, Algorithm \ref{alg:fas}.  This FAS V-cycle appears in various places, for example Algorithm 14 of \cite{Bruneetal2015} or in section 5.3.4 of \cite{Trottenbergetal2001}.

\begin{pseudofloat}[ht]
\begin{pseudo} \label{ps:fas-vcycle}
\pr{fas-vcycle}(\ell^J; u^J)\text{:} \\+
    for $j=J$ downto $j=1$ \\+
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j;u^j)$ \\
      $u^{j-1} = \iR u^j$ \\
      $\ell^{j-1} = f^{j-1}(u^{j-1}) + R \left(\ell^j - f^j(u^j)\right)$ \\-
    $\text{\pr{solve}}(\ell^0;u^0)$ \\
    for $j=1$ to $j=J$ \\+
      $u^j \gets u^j + P (u^{j-1} - \iR u^j)$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j;u^j)$ \\-
\end{pseudo}
\caption{The standard FAS V-cycle for a discretized PDE $f^J(u^J)=\ell^J$.}
\label{alg:fas}
\end{pseudofloat}

@@

FIXME remove this appendix, after using it in the introduction and discussion etc.
